SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-05 design_impl.tex
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

\subsection{Changes made to original SPP}
\label{Impl-Changes}
To modify the SPP design to suit our scheme, the following changes
were made:

\begin{itemize}
\item \textbf{Original Thresholds discarded}\newline In PPF, the
  perceptron sum is used to decide whether to prefetch or not, and the
  fill-level in case of prefetch.  Thus, the confidence thresholds --
  T\textsubscript{f} and T\textsubscript{p} are no longer needed.

% EB: Can this point be removed?
% djimenez: I think this would be necessary to reproduce our results so leave it in. 
% EB: It's ultimately a check to avoid SPP's internal segfault. Won't have a real impact.

%\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The
%  original SPP does not consider the slots available in the L2 MSHR
%  queue.  It has the assumption that at no time can the number of
%  suggested prefetches exceed the capacity of the queue.  While it was
%  not described in the paper, the internal confidence mechanism makes
%  sure that the above assumption was maintained.  In PPF, since there
%  cannot be an assumption on the degree of aggressiveness of the
%  underlying prefetcher, we explicitly check that at no point should
%  the number of suggested prefetches exceed the L2 MSHR queue. \pg{is
%    this a big deal?  Seems like a detail}

\item \textbf{Enhanced Prefetch Table}\newline The prefetch table
  introduced in the original SPP keeps track of up to 1024 L2 cache
  prefetch suggestions.  For PPF, the prefetch table is modified to
  store the features used for perceptron indexing..
  This metadata is needed to retrieve the
  values of the features to train the perceptron at a later stage in
  time. The prefetch table mentioned above keeps track of the
  prefetches that actually happened in the L2 Cache.  We introduce the
  reject table to keep a track of the prefetches that were suggested
  by the base SPP engine but were rejected by PPF.  In case it is
  detected that a future demand fetch could have used this rejected
  prefetch, the perceptron weights are updated to reflect that
  misprediction.
\end{itemize}

\subsection{Features used by Perceptron}
\label{Impl-Features}
Here we discuss the various features that correlate the prefetching
decision with the program behavior. All the features we used can be 
derived from the information available in the L2 Cache access stream.
Our feature selection involved searching over a large space of 
relevant perceptron features. Using the statistical methodology 
outlined in Section~\ref{Method-Features}, we pruned the feature 
set to a minimal yet relevant set of features
% \pg{again you want to get at why
% these were chosen.  Also mention that you searched over a large
% space of possible features and these were the best.}

% [EB] Tried addressing. Does it look redundant when read after 'Feature Selection'
% bullet from the previous section?

\begin{itemize}
\item \textbf{Base Address}: The base address of the demand 
access that triggers the prefetch corresponds to a stream of 
future accesses that SPP and the PPF have seen before. 
Therefore, PPF will correlate the past behavior of this address 
to the confidence of a prefetch.

\item \textbf{Cache Line} and \textbf{Page Address}: These two
  separate features are derived from the base address that triggered
  the prefetch, as follows: base\_addr >> LOG2\_BLOCK \_SIZE and
  base\_addr >> LOG2\_PAGE\_SIZE respectively. The idea behind using
  three different shifted versions of the same feature is that it
  allows us to look into a wider range of bits than with a single
  version.  It also helps give more importance to the overlapping bits
  and lesser importance to most and least significance bits. This
  approach can also eliminate destructive interference that can be
  caused by directly folding the address bits into half.

% djimenez: this is a nice intuition and explanation. 
% [EB] Thank you!

\item \textbf{Program Counter XOR Depth}: The PC is for the
  instruction that triggered the prefetch chain.  Depth refers to the
  iteration count of the lookahead stages.  In general, the PC is not
  considered as a good basis for doing lookahead prefetching as all
  the prefetches with depth >= 1 are aliased into the same PC which
  is not true in an actual demand access.  This feature resolves
  a PC into a different value for each lookahead depth of prefetch
  speculation, giving a more accurate correlation in lookahead cases.
  This is akin to the concept of Virtual Program Counters~\cite{VPC}
  introduced by Kim \textit{et. al.} for indirect branch prediction.

\item \textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR
    PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$
  PC before the instruction that triggered the current prefetch.
  Hashing together the last three PCs tell PPF about the path that led
  to the current demand access and helps capture and branching
  information of the current basic block.  PCs are shifted in the
  increasing order of history before being hashed together.  This is
  done to avoid the resultant value of zero when 2 or more PCs are the
  same.  Additionally, obfuscating the information as it gets old
  allows us to get a wider and approximate look into the program's
  history.

\item \textbf{Program Counter XOR Delta}: This feature tells us if a
  given PC favors particular value(s) of delta.  As noted earlier,
  while the PC alone does not convey useful information, this
  hash resolves the PC into different values based on the tendency of
  that PC to favor a certain delta.  Thus, the dynamic nature of
  different instances of the same memory access instruction is 
  captured here.
 % GAC: Seems more predictable than dynamic if its capturing 
 % what is essentially a streaming access pattern. Not quite sure how to rephrase,
 % if I'm correct

 % GAC: Again having trouble phrasing the below point. I was attempting to say
 % that despite SPP having low confidence due to speculation, PPF may 
 % reinforce prefetches with lower confidence if SPP has had a tendency to be 
 % correct in the past.
\item \textbf{Confidence}: The integer confidence, on a scale of 0 to
  100, that was used in the original SPP design. While the original 
  confidence does not directly make the decision to prefetch,
  PPF correlates it to the correctness of a proposed prefetch.
  While the original SPP may have dismissed a prefetch due to running
  further into speculation, PPF can use the original confidence as indicator
  of not only when prefetches become less confident, but also how likely
  a low confident speculation is correct in the context of other features.

\item \textbf{Page Address XOR Confidence}: This feature scores the
  tendency of each page to be prefetch friendly or prefetch averse. It
  helps resolve a page into different entries depending on its
  confidence for prefetching, which can vary during phases of a
  program execution.

\item \textbf{Current Signature XOR Delta}: Recall from the discussion
  of SPP in Section \ref{Background-SPP} that the new signature is
  generated using the old signature and the current block delta.
%Gino: Added the sentence below, is this correct?
%[EB]: It gives the next signature instead of next delta
  The result of this feature is the next signature that is predicted to be
  accessed based on the delta predicted by SPP. While ``Current
  Signature XOR Delta'' is not the exact formula for generating the
  future signature, it gives an approximate idea of the path that the
  combination of these two values can lead to.

%%%%%%% Below are the features rejected when downsizing 14 -> 9
% \item \textbf{Current Signature}: The 12-bit signature used by SPP
%   to index into the Pattern Table
% \item \textbf{Delta}: The signed version of the difference in the
%   block address that triggers the prefetch and the block address of
%   the data being prefetched
% \item \textbf{PC}: Program Counter of the Load instruction that
%   triggers that particular prefetch
% \item \textbf{Base address XOR Delta}: If any particular value(s) of
%   delta are highly favored by a given address which triggered the
%   prefetch, this composite feature can capture that information
% \item \textbf{Cache line XOR Delta}: this feature captures if any
%   particular delta is highly favored by a given cache line address
%\item \textbf{Confidence XOR Delta}: This gives a combined insight
%  into that fact that certain delta-confidence pairs predicted by
%  SPP might correlate more with the prefetch outcome
%%%%%%%%%%%%%%%%%%%%%%

\end{itemize}

Some of the composite features are derived from simple hashing (XOR)
of two primary features.  There is always a question of usefulness of
such composite features and the new information added.  We justify the
choice of each feature by quantifying the contribution made towards
predicting prefetch behavior, in Section \ref{Method-Features}.
Finally, as noted above, each feature indexes into its independent
entry of perceptron weights.

>>>>

\subsection{Changes Made to SPP}
\label{Impl-Changes}
To modify the SPP design to suit our scheme, the following changes were made: \\ \\
%
\textbf{Original Thresholds Discarded:} In PPF, the perceptron sum is used to
decide whether to prefetch or not, and the fill-level in case of prefetch.
Thus, the confidence thresholds -- T\textsubscript{f} and T\textsubscript{p}
are no longer needed. \\
% djimenez: keeping comments on every line here so it doesn't screw up my
%"roll-your-own" itemized list. don't put in any newlines.
%
%
% EB: Can this point be removed?
% djimenez: I think this would be necessary to reproduce our results so leave it in. 
% EB: It's ultimately a check to avoid SPP's internal segfault. Won't have a real impact.
%
%\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The
%  original SPP does not consider the slots available in the L2 MSHR
%  queue.  It has the assumption that at no time can the number of
%  suggested prefetches exceed the capacity of the queue.  While it was
%  not described in the paper, the internal confidence mechanism makes
%  sure that the above assumption was maintained.  In PPF, since there
%  cannot be an assumption on the degree of aggressiveness of the
%  underlying prefetcher, we explicitly check that at no point should
%  the number of suggested prefetches exceed the L2 MSHR queue. \pg{is
%    this a big deal?  Seems like a detail}
%
\textbf{Enhanced Prefetch Table:} The prefetch table introduced in the
original SPP keeps track of up to 1024 L2 cache prefetch suggestions.  For
PPF, the prefetch table is modified to store the features used for perceptron
indexing..  This metadata is needed to retrieve the values of the features to
train the perceptron at a later stage in time. The prefetch table mentioned
above keeps track of the prefetches that actually happened in the L2 Cache.
We introduce the reject table to keep a track of the prefetches that were
suggested by the base SPP engine but were rejected by PPF.  In case it is
detected that a future demand fetch could have used this rejected prefetch,
the perceptron weights are updated to reflect that misprediction.

\subsection{Features used by Perceptron}
\label{Impl-Features}
Here we discuss the various features that correlate the prefetching
decision with the program behavior. All the features we used can be 
derived from the information available in the L2 Cache access stream.
Our feature selection involved searching over a large space of 
relevant perceptron features. Using the statistical methodology 
outlined in Section~\ref{Method-Features}, we pruned the feature 
set to a minimal yet relevant set of features. \\ \\
% \pg{again you want to get at why
% these were chosen.  Also mention that you searched over a large
% space of possible features and these were the best.}
%
% [EB] Tried addressing. Does it look redundant when read after 'Feature Selection'
% bullet from the previous section?
%
\textbf{Base Address}: The base address of the demand access that triggers the
prefetch corresponds to a stream of future accesses that SPP and the PPF have
seen before.  Therefore, PPF will correlate the past behavior of this address
to the confidence of a prefetch. \\
%
\textbf{Cache Line} and \textbf{Page Address}: These two separate features are
derived from the base address that triggered the prefetch, as follows:
base\_addr >> LOG2\_BLOCK \_SIZE and base\_addr >> LOG2\_PAGE\_SIZE
respectively. The idea behind using three different shifted versions of the
same feature is that it allows us to look into a wider range of bits than with
a single version.  It also helps give more importance to the overlapping bits
and lesser importance to most and least significance bits. This approach can
also eliminate destructive interference that can be caused by directly folding
the address bits into half. \\
%
% djimenez: this is a nice intuition and explanation. 
% [EB] Thank you!
\textbf{Program Counter XOR Depth}: The PC is for the instruction that
triggered the prefetch chain.  Depth refers to the iteration count of the
lookahead stages.  In general, the PC is not considered as a good basis for
doing lookahead prefetching as all the prefetches with depth >= 1 are aliased
into the same PC which is not true in an actual demand access.  This feature
resolves a PC into a different value for each lookahead depth of prefetch
speculation, giving a more accurate correlation in lookahead cases.  This is
akin to the concept of Virtual Program Counters~\cite{VPC} introduced by Kim
\textit{et. al.} for indirect branch prediction. \\
%
\textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR
PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$
PC before the instruction that triggered the current prefetch.
Hashing together the last three PCs tell PPF about the path that led
to the current demand access and helps capture and branching
information of the current basic block.  PCs are shifted in the
increasing order of history before being hashed together.  This is
done to avoid the resultant value of zero when 2 or more PCs are the
same.  Additionally, blurring the information as it gets older
allows us to get a wider and approximate look into the program's
history. \\
%
\textbf{Program Counter XOR Delta}: This feature tells us if a given PC favors
particular value(s) of delta.  As noted earlier, while the PC alone does not
convey useful information, this hash resolves the PC into different values
based on the tendency of that PC to favor a certain delta.  Thus, the dynamic
nature of different instances of the same memory access instruction is
captured here. \\
%
 % GAC: Seems more predictable than dynamic if its capturing 
 % what is essentially a streaming access pattern. Not quite sure how to rephrase,
 % if I'm correct
%
 % GAC: Again having trouble phrasing the below point. I was attempting to say
 % that despite SPP having low confidence due to speculation, PPF may 
 % reinforce prefetches with lower confidence if SPP has had a tendency to be 
 % correct in the past.
\textbf{Confidence}: The integer confidence, on a scale of 0 to 100, that was
used in the original SPP design. While the original confidence does not
directly make the decision to prefetch, PPF correlates it to the correctness
of a proposed prefetch.  While the original SPP may have dismissed a prefetch
due to running further into speculation, PPF can use the original confidence
as indicator of not only when prefetches become less confident, but also how
likely a low confident speculation is correct in the context of other
features. \\
\textbf{Page Address XOR Confidence}: This feature scores the tendency of each
page to be prefetch friendly or prefetch averse. It helps resolve a page into
different entries depending on its confidence for prefetching, which can vary
during phases of a program execution. \\
%
\textbf{Current Signature XOR Delta}: Recall from the discussion of SPP in
Section \ref{Background-SPP} that the new signature is generated using the old
signature and the current block delta.  The result of this feature is the next
signature that is predicted to be accessed based on the delta predicted by
SPP. While ``Current Signature XOR Delta'' is not the exact formula for
generating the future signature, it gives an approximate idea of the path that
the combination of these two values can lead to. \\ \\
%
%Gino: Added the sentence below, is this correct?
%[EB]: It gives the next signature instead of next delta
%
%%%%%%% Below are the features rejected when downsizing 14 -> 9
% \item \textbf{Current Signature}: The 12-bit signature used by SPP
%   to index into the Pattern Table
% \item \textbf{Delta}: The signed version of the difference in the
%   block address that triggers the prefetch and the block address of
%   the data being prefetched
% \item \textbf{PC}: Program Counter of the Load instruction that
%   triggers that particular prefetch
% \item \textbf{Base address XOR Delta}: If any particular value(s) of
%   delta are highly favored by a given address which triggered the
%   prefetch, this composite feature can capture that information
% \item \textbf{Cache line XOR Delta}: this feature captures if any
%   particular delta is highly favored by a given cache line address
%\item \textbf{Confidence XOR Delta}: This gives a combined insight
%  into that fact that certain delta-confidence pairs predicted by
%  SPP might correlate more with the prefetch outcome
%%%%%%%%%%%%%%%%%%%%%%
%
Some of the composite features are derived from simple hashing (XOR)
of two primary features.  There is always a question of usefulness of
such composite features and the new information added.  We justify the
choice of each feature by quantifying the contribution made towards
predicting prefetch behavior, in Section \ref{Method-Features}.
Finally, as noted above, each feature indexes into its independent
entry of perceptron weights.

### 2018-12-05 intro.tex
Adding Perc image, motivation section
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

Figure~\ref{Fig:Motivation} illustrates the above scenario in a working manner.
Here we consider a state-of-the-art lookahead prefetcher -- SPP. We keep on 
re-tuning its thresholds, each time getting a variant with an increasing lookahead 
depth coverage. The figure depicts the behavior of the prefetcher variants running 
on the trace \textit{603.bwaves\_s-1740B}. The IPC, total number of prefetches 
suggested by the prefetcher (TOTAL\_PF) and the actual useful predictions (GOOD\_PF),
all have been normalized to the least aggressive version of the prefetcher. 
As the look-ahead depth of the prefetch suggestions increase, so do useful prefetches. 
But that comes at a cost of total prefetches increasing at an even higher rate. This leads 
to cache pollution and bandwidth contention, as is clearly visible with the reduction in 
the IPC. Similar behavior was observed in 12 other SPEC CPU 2017 simpoints.

@@

\vspace{1ex}This paper describes PPF, explains its merits, offers analysis, and
outlines the scope for future research. Its contributions include:

@@

\item An on-line neural model used for hardware data
  prefetching.  Previous work in this area either relied on program
  semantics~\cite{Semantics} or were application specific~\cite{Datacenter}.
% djimenez: removing reference to this arXiv paper. we are not obligated to cite it.
%, or
%  could not yield a significant improvement over the baseline prefetcher
%  used~\cite{BadPerc}.

\item Implementing PPF in a robust and accurate
  practical prefetching mechanism, yielding a significant performance
  improvement compared to other state-of-the-art prefetchers.
  Moreover, PPF learns to adapt itself to shared resource
  constraints, leading to further increased performance in multi-core and
  bandwidth-constrained environments.

@@

The paper is organized as follows. Section~\ref{Background} describes
background work and summarizes SPP.  Section~\ref{Arch} gives the 
architectural overview of PPF. Section~\ref{Impl} discusses the
implementation of PPF using SPP and the particular features used.
Section~\ref{Method} gives the evaluation methodology and explores the feature
space for perceptron learning.  Section~\ref{Results} evaluates the
performance of the prefetcher and Section~\ref{Conclusion} concludes the
paper.

>>>>

Figure~\ref{Fig:Motivation} illustrates the above scenario in a working manner.
Here we consider a state-of-the-art lookahead prefetcher -- SPP. We keep on 
re-tuning its thresholds, each time getting a variant with an increasing lookahead 
depth coverage. The figure depicts the behavior of the prefetcher variants running 
on the trace {\tt 603.bwaves\_s-1740B}. The IPC, total number of prefetches 
suggested by the prefetcher (TOTAL\_PF) and the actual useful predictions (GOOD\_PF),
all have been normalized to the least aggressive version of the prefetcher. 
As the look-ahead depth of the prefetch suggestions increase, so do useful prefetches. 
But that comes at a cost of total prefetches increasing at an even higher rate. This leads 
to cache pollution and bandwidth contention, as is clearly visible with the reduction in 
the IPC. Similar behavior was observed in 12 other SPEC CPU 2017 simpoints.

@@

This paper describes PPF, explains its merits, offers analysis, and
outlines the scope for future research. Its contributions are:

@@

\item An on-line neural model used for hardware data prefetching.  Previous
work in this area either relied on program semantics~\cite{Semantics} or were
application specific~\cite{Datacenter}.

\item Implementing PPF filtering a state-of-the-art prefetchier, giving a
significant performance improvement compared to previous work.  PPF learns to
adapt itself to shared resource constraints, leading to further increased
performance in multi-core and bandwidth-constrained environments.

@@

% djimenez - cutting this
%The paper is organized as follows. Section~\ref{Background} describes
%background work and summarizes SPP.  Section~\ref{Arch} gives the 
%architectural overview of PPF. Section~\ref{Impl} discusses the
%implementation of PPF using SPP and the particular features used.
%Section~\ref{Method} gives the evaluation methodology and explores the feature
%space for perceptron learning.  Section~\ref{Results} evaluates the
%performance of the prefetcher and Section~\ref{Conclusion} concludes the
%paper.

### 2018-12-05 method.tex
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

The block size is fixed at 64 bytes. Prefetching is only triggered at an L2
cache demand accesses but could be directed to the L2 or last-level cache. No
L1 data level prefetching is done. The LRU replacement policy is used on all
levels of cache hierarchies. Branch prediction is done using the perceptron
branch predictor~\cite{Perc_Branch}.  The page size is 4KB.  ChampSim operates
all the prefetchers strictly in the physical address space.

\subsection{Testing Under Additional Memory Constraints}
\label{Method-AdditionalMem}
The default single-core configuration simulates a 2MB LLC and a single
channel DRAM with 12.8GB/s bandwidth.  We extend the simulations to
include memory constraints introduced in DPC-2.  Specifically we look
at the following two variations:
\begin{itemize}
\item \textit{Low Bandwidth DRAM}: Here the DRAM bandwidth is limited
  to 3.2 GB/s
\item \textit{Small LLC}: In this scenario, LLC size is reduced to 512
  KB
\end{itemize}
All the multi-core simulations are only done in the default
configuration.

@@

\subsection{Developing Features for PPF}
\label{Method-Features}
This section describes the intuition and analysis that went behind developing
the features.  As noted earlier, we developed a set of nine features that allow
the perceptron layer to correlate prefetching decision with the program
context.  To study the correlation across each feature, we study statistically
the perceptron weights and try to interpret their distribution.

\textbf{Global Pearson's Correlation}\newline For this experiment, we create a
dump of the state of perceptron weights at the end of all trace execution.
The weights obtained from running all the SPEC CPU 2017 traces are concatenated.
Since the weights are collected at the end of individual trace execution, the
perceptron weights have attained a relatively stable value by now.  Hence,
this dump represents a `snap' of the trained perceptron weights across all the
SPEC CPU 2017 traces.  The intuition here is that the feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome.  On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

We plot a histogram for each feature depicting weights distribution
from -16 to +15 and use this histogram to generate the Pearson's
correlation factor for that feature.  In statistics, Pearson's factor
is a numerical measure of the degree of linear correlation between two
variables.  It ranges from -1 to +1.  The magnitude of Pearson's
factor depicts the extent of correlation and the sign depicts whether
it is a positive correlation or a negative correlation.  Values close
to 0 suggest a low correlation or at times, noisy data.  A value of
+1/-1 suggests a perfectly linear positive / negative correlation
respectively.  Figure~\ref{Fig:Hist} depicts the histogram distribution 
of trained weights, for a feature with high and another with moderate
Pearson's coefficient.

Figure~\ref{Fig:Pvals} shows all the features used, arranged in the increasing
order of their Pearson's factor.  As can be seen 5 out of the 9
features provide a moderate to high correlation, with the magnitude of
P-value > 0.6.  The single most important feature, 
\textit{Confidence XOR Page address} helps provide
a correlation to prefetch outcome with a factor of 0.90.

\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces.  Here we give special attention to features
with low P-values in.  Figure~\ref{Pvals_Traces} shows the
variation of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and 
\textit{PC XOR Depth}; across all the SPEC CPU 2017 traces.  
For simplicity, the traces are arranged in an
increasing order of contribution made by the feature and the trace
names have been omitted.  It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.5) for
a significant number of traces.

\textbf{Trimming Features Using Correlation} \newline Besides
providing interesting insights into prefetching behavior, P-value can
also be used for feature selection and prefetcher tuning.

Here, we introduce the concept of cross-correlation across features.
Just as we examined correlation of each feature with the final
outcome, we can also study correlation between the features.  We used
the above methodology to eliminate features that provide a little or
no information that has already been captured in other features.

As a part of feature selection, we initially came up a mix of 23
features- primary and composite.  By studying cross correlation of
each of these features against others in a 23x23 matrix, we identified
pairs of features with correlation factor > 0.9 in magnitude and
carefully eliminated redundant features.  Using this approach, we
managed to reduce the feature count to 9.  Thus, in the final
implementation of PPF, no two features have a high correlation
between them.  This way we can be sure that each feature is bringing
in contribution that cannot be captured using other features.

Secondly, studying the relative importance of each feature enabled 
us to vary the number of entries dedicated for each feature.  
Features with higher Pearson's correlation, \textit{cache line} 
and \textit{page address}  were given most importance and
allowed full 12-bits of indexing.  Features like \textit{IP XOR delta} 
and \textit{IP XOR depth} with a low overall P-value 
were allocated fewer indexes in the feature table.

To conclude, in the above discussion we justify the features for
perceptron from a statistical viewpoint.  We also show how this
information can also be used for prefetcher tuning.  All this study
was made possible only because we used on-line perceptron learning for
prefetching and that enabled us to examine the weights in detail.

@@

\begin{table}[h]
\begin{adjustwidth}{-0.1cm}{}
    \centering
    \begin{tabular}{|c|c|c|c|}
    \hline
        \textbf{Structure} &
        \textbf{Entry} &
        \textbf{Components} &
        \textbf{Total} \\
    \hline
                                            &  \multirow{5}{0.5cm}{256}    & Valid (1 bit)  &             \\
                                             &      & Tag (16 bits)        &  \multirow{2}{0.9cm}{11008}           \\
                            Signature Table  &   & Last Offset (6 bits) &  \multirow{2}{0.5cm}{bits}  \\  
                                             &      & Signature (12 bits)  &             \\
                                             &      & LRU (6 bits)         &             \\
    \hline
                                    &  \multirow{3}{0.5cm}{512}    & $C_{sig}$ (4bits)      &\multirow{2}{0.9cm}{24576}               \\
                       Pattern Table         &   & $C_{delta}$ (4*4 bits) &  \multirow{2}{0.5cm}{bits}  \\
                                             &      & Delta (4*7 bits)       &               \\
    \hline
        \multirow{4}{1.5cm}{Perceptron\newline}     & 4096*4    & \multirow{4}{0.8cm}{5 bits}  & \multirow{3}{1.1cm}{113280}             \\
        \multirow{3}{1.2cm}{Weights}                & 2048*2    &           &  \multirow{3}{0.5cm}{bits}  \\
                                                    & 1024*2    &           &               \\
                                                    & 128*1     &           &              \\
    \hline
        Prefetch                & \multirow{2}{0.7cm}{1024}      & \multirow{2}{1cm}{85 bits}       & 87040 \\
        Table\footnotemark[1]   &           &               & bits\\
    \hline
        Reject                & \multirow{2}{0.7cm}{1024}      & \multirow{2}{1cm}{84 bits}    & 86016 \\
        Table\footnotemark[2] & & & bits\\
    \hline
        \multirow{4}{1.0cm}{Global\newline\newline}   & \multirow{4}{0.2cm}{8} & Signature (12 bits)  & \multirow{4}{1.1cm}{264 bits} \\
        \multirow{3}{1.1cm}{History\newline}        &                        & Confidence (8 bits)  &                               \\
        \multirow{2}{1.2cm}{Register}               &                        & Last Offset (6 bits) &                               \\
                                                    &                        & Delta (7 bits)       &                               \\
    \hline
        Accuracy        & 1     & C$_{total}$       & 10 bits   \\
        Counters        & 1     & C$_{useful}$      & 10 bits   \\
    \hline
        \multirow{3}{1.5cm}{Global PC\newline}      &       & $PC_1$ (12 bits)      &           \\
        \multirow{2}{1.5cm}{~Trackers}              & 3     & $PC_2$ (12 bits)      & 36 bits   \\
                                                    &       & $PC_3$ (12 bits)      &           \\
    \hline
        \multicolumn{4}{|c|}{\textbf{Total: 3222940 bits = 39.34 KB}}\\
    \hline
    \end{tabular}
    \caption{SPP-Perc Storage Overhead}
    \label{tab:PPF_overhead}
\end{adjustwidth}
\end{table}

>>>>

The block size is fixed at 64 bytes. Prefetching is only triggered at an L2
cache demand accesses but could be directed to the L2 or last-level cache. No
L1 data level prefetching is done. The LRU replacement policy is used on all
levels of cache hierarchies. Branch prediction is done using the perceptron
branch predictor~\cite{PerceptronPredictor}.  The page size is 4KB.  ChampSim operates
all the prefetchers strictly in the physical address space.

\subsection{Testing Under Additional Memory Constraints}
\label{Method-AdditionalMem}
The default single-core configuration simulates a 2MB LLC and a single channel
DRAM with 12.8GB/s bandwidth.  We extend the simulations to include memory
constraints introduced in DPC-2.  Specifically we look at the following two
variations: Low Bandwidth DRAM, where DRAM bandwidth is limited to 3.2 GB/s,
and small LLC, where the LLC size is reduced to 512 KB All the multi-core
simulations are only done in the default configuration.

@@

\subsection{Developing Features for PPF}
\label{Method-Features}
This section describes the intuition and analysis that went behind developing
the features.  As noted earlier, we developed a set of nine features that allow
the perceptron layer to correlate prefetching decision with the program
context.  To study the correlation across each feature, we study statistically
the perceptron weights and try to interpret their distribution. \\ \\
%
\textbf{Global Pearson's Correlation}\newline For this experiment, we examine
the perceptron weights at the end of all trace execution by which time the
weights have settled to steady values.  The weights obtained from running all
the SPEC CPU 2017 traces are concatenated.  The feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome.  On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

We plot a histogram for each feature depicting weights distribution from -16
to +15 and generate the Pearson's correlation factor for that feature.
Pearson's factor is a numerical measure ranging from -1 to 1 of the degree of
linear correlation between two variables.  The magnitude of Pearson's factor
gives the extent of correlation and the sign indicates whether it is a
positive correlation or a negative correlation.  Values close to 0 suggest a
low correlation or at times, noisy data.  A value of +1/-1 suggests a
perfectly linear positive / negative correlation respectively.
Figure~\ref{Fig:Hist} depicts the histogram distribution of trained weights,
for a feature with high and another with moderate Pearson's coefficient.

Figure~\ref{Fig:Pvals} shows all the features used, arranged in the increasing
order of their Pearson's factor.  As can be seen 5 out of the 9
features provide a moderate to high correlation, with the magnitude of
P-value > 0.6.  The single most important feature, 
\textit{Confidence XOR Page address} helps provide
a correlation to prefetch outcome with a factor of 0.90. \\ \\
%
\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces.  Here we give special attention to features
with low P-values in.  Figure~\ref{Pvals_Traces} shows the
variation of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and 
\textit{PC XOR Depth}; across all the SPEC CPU 2017 traces.  
For simplicity, the traces are arranged in an
increasing order of contribution made by the feature and the trace
names have been omitted.  It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.5) for
a significant number of traces. \\ \\
%
\textbf{Trimming Features Using Correlation} \newline Beside
providing interesting insights into prefetching behavior, P-value can
also be used for feature selection and prefetcher tuning.

As we examine correlation of each feature with the final outcome, we also
study correlation between the features.  We used the above methodology to
eliminate features providing little information that has already been captured
in other features.

We initially came up a mix of 23 features- primary and composite.  By studying
cross correlation of each of these features against others in a 23$\times$23
matrix, we identified pairs of features with correlation factor > 0.9 in
magnitude and eliminated redundant features, reducing the feature count to 9.
Thus, in the final implementation of PPF, no two features have high
correlation between them.  This way we can be sure that each feature makes a
contribution that cannot be captured using other features.

Secondly, studying the relative importance of each feature enabled us to vary
the number of entries dedicated for each feature.  Features with higher
correlation, \textit{cache line} and \textit{page address} were given most
importance and allowed full 12-bits of indexing.  Features like \textit{IP XOR
delta} and \textit{IP XOR depth} with a low overall P-value were allocated
fewer entries in the feature table.

%To conclude, in the above discussion we justify the features for
%perceptron from a statistical viewpoint.  We also show how this
%information can also be used for prefetcher tuning.  All this study
%was made possible only because we used on-line perceptron learning for
%prefetching and that enabled us to examine the weights in detail.

@@

\begin{table}[h]
\begin{adjustwidth}{-0.1cm}{}
    \centering
    \begin{tabular}{|c|c|c|c|}
    \hline
        \textbf{Structure} &
        \textbf{Entry} &
        \textbf{Components} &
        \textbf{Total} \\
    \hline
                                            &  \multirow{5}{0.5cm}{256}    & Valid (1 bit)  &             \\
                                             &      & Tag (16 bits)        &  \multirow{2}{0.9cm}{11008}           \\
                            Signature Table  &   & Last Offset (6 bits) &  \multirow{2}{0.5cm}{bits}  \\  
                                             &      & Signature (12 bits)  &             \\
                                             &      & LRU (6 bits)         &             \\
    \hline
                                    &  \multirow{3}{0.5cm}{512}    & $C_{sig}$ (4bits)      &\multirow{2}{0.9cm}{24576}               \\
                       Pattern Table         &   & $C_{delta}$ (4*4 bits) &  \multirow{2}{0.5cm}{bits}  \\
                                             &      & Delta (4*7 bits)       &               \\
    \hline
        \multirow{4}{1.5cm}{Perceptron\newline}     & 4096*4    & \multirow{4}{0.8cm}{5 bits}  & \multirow{3}{1.1cm}{113280}             \\
        \multirow{3}{1.2cm}{Weights}                & 2048*2    &           &  \multirow{3}{0.5cm}{bits}  \\
                                                    & 1024*2    &           &               \\
                                                    & 128*1     &           &              \\
    \hline
        Prefetch                & \multirow{2}{0.7cm}{1024}      & \multirow{2}{1cm}{85 bits}       & 87040 \\
        Table\footnotemark[1]   &           &               & bits\\
    \hline
        Reject                & \multirow{2}{0.7cm}{1024}      & \multirow{2}{1cm}{84 bits}    & 86016 \\
        Table\footnotemark[2] & & & bits\\
    \hline
        \multirow{4}{1.0cm}{Global\newline\newline}   & \multirow{4}{0.2cm}{8} & Signature (12 bits)  & \multirow{4}{1.1cm}{264 bits} \\
        \multirow{3}{1.1cm}{History\newline}        &                        & Confidence (8 bits)  &                               \\
        \multirow{2}{1.2cm}{Register}               &                        & Last Offset (6 bits) &                               \\
                                                    &                        & Delta (7 bits)       &                               \\
    \hline
        Accuracy        & 1     & C$_{total}$       & 10 bits   \\
        Counters        & 1     & C$_{useful}$      & 10 bits   \\
    \hline
        \multirow{3}{1.5cm}{Global PC\newline}      &       & $PC_1$ (12 bits)      &           \\
        \multirow{2}{1.5cm}{~Trackers}              & 3     & $PC_2$ (12 bits)      & 36 bits   \\
                                                    &       & $PC_3$ (12 bits)      &           \\
    \hline
        \multicolumn{4}{|c|}{\textbf{Total: 3,222,940 bits = 39.34 KB}}\\
    \hline
    \end{tabular}
    \caption{SPP-Perc Storage Overhead}
    \label{tab:PPF_overhead}
\end{adjustwidth}
\end{table}
